### 2014-08-12 Studies/2011_San_Francisco/Report/San_Francisco_Fire.tex
San Francisco: Edits to narrative and incident timeline

The account of events for this incident was documented by the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2011-13~\cite{NIOSH:Bowyer2}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer2}:
\begin{quote}
At approximately 1045 hours, 3 engine companies, 2 truck companies, 1 heavy rescue squad, and 3 battalion chiefs were dispatched to a report of curtains on fire due to an electrical short at a single family residential structure. E26, with [the lieutenant] and [the fire fighter/paramedic], arrived at approximately 1048 hours, positioned just north of the front door on [the front side] of the fire structure on the narrow street. [The lieutenant] initially grabbed the CO$_2$ extinguisher because of a reported electrical fire. After consulting with the occupant, [the lieutenant] and [the fire fighter/paramedic] pulled a 1~3/4'' hoseline off the engine and made entry through the front door ([front side] street level), noticing light smoke. [The lieutenant] radioed on the fireground channel that ``smoke was showing and we have a working fire below grade.'' The transmission was acknowledged by both BC6 and DC 3.

E24 arrived at approximately 1051 hours and positioned down the hill to allow T11, who was still en route, to get around them. At approximately 1052 hours, E11 and BC6 (shift relief BC with no assigned aide), traveling north with a view of [the rear side] of the structure, noticed no smoke or fire visible. E11 caught the hydrant just north of the structure and laid 3'' supply lines to both sides of E26. BC6 drove around E11 and parked in a neighboring driveway just north of the fire structure. BC6 (initial IC) radioed E26 for their location and they replied they were trying to locate the fire. The IC walked to the front of the fire structure and radioed E26 again about their status but got no response. The E11 crew was in the process of building a wye, when the IC instructed them to get a 1~3/4'' hoseline and some irons and go down the exterior wooden stairway of [the left side of the structure] and make entry at the door that was below grade. At approximately 1053 hours, BC9 arrived on scene and was assigned to supervise fire attack group responsibilities per protocol as second due BC. BC9 followed the E26's hoseline through the front door and encountered the E26 crew ([the lieutenant and the fire fighter/paramedic]) operating in the front hallway on the street level floor near the entrance to the garage and stairway to the basement. [The lieutenant] stated that the fire must be a floor below them. BC9 stated they would attack the fire from the [left side] of the structure and exited the front door, but the [lieutenant and the fire fighter/paramedic] did not follow. At approximately 1054 hours, the E24 crew made entry into the garage to check for access and attempt to intercept the E26 crew. They thought they heard what could have been SCBA cylinders clanging and hitting the floor just inside the doorway.

At approximately 1056 hours, T15 and DC3 arrived on scene. DC3 took command as IC. The IC assigned T15 to the roof for ventilation. RS1 signaled to command they were going to split their crew per their normal protocol while operating with an engine company: 2 crew members going on first floor (RS1A) and 2 crew members going to the second floor (RS1B). The IC acknowledged. The RS1A entered the garage and headed for the door in the right corner of the garage to meet up with E24, but struggled since the door was blocked by vehicles and storage boxes inside the garage. The IC did a face-to-face with BC6 and instructed BC6 to get a layout of the fire structure from a resident.

At approximately 1057 hours, BC9 went down a debris filled path along the outside of the [left side of the structure] to the wooden stairs connecting the basement door to the sub-basement door. He went to the sub-basement and forced the door open (see Photo 1). After seeing the room filled with light smoke, he looked around the B/C corner and saw smoke but no fire. BC6 located an older woman on the street, who resided at the fire structure and tried to get an understanding of the layout. At this point, dark smoke was observed about 2 feet off the ceiling in the garage. The smoke was not turbulent. At approximately 1058 hours, BC9 headed to the basement level man door. The fire self-vented out the glass door near the C/D corner (see Photo 2). BC9 forced the barred security door open with his halligan bar and then kicked the door open. The E11 crew advanced into the basement with the 1~3/4'' hoseline. At the door, they felt a blast of heat and observed heavy smoke in the fully involved room. The heat forced the E11 fire fighter operating the nozzle to back up. BC9 radioed, ``We have heavy fire and smoke and need a second line on [the left side of the structure].'' (see Photo 3).

E32 had arrived on scene and reported to ``Command'' minutes earlier as RIC but the IC reassigned them to back up E11 on [the left side of the structure]. The IC requested another RIC company from dispatch. At approximately 1100 hours, the E24 crew and RS1A were in the garage trying to make entry into the house but were unable to get past the doorway due to the extreme heat (see Photo 4). BC6 went to the IC to explain the layout and had noticed the dramatic change in the smoke (i.e., from gray to dark in color, lofting to pressurized speed, and medium to heavy volume) coming out of the garage covering the command post in the street. The IC tried again to radio E26 but received no response. The IC radioed E26 again and heard a garbled response. The IC thought the E26 crew was on the hoseline and near BC9. At approximately 1102 hours, the E32 crew had deployed a charged 1~3/4'' hoseline down to [the left side of the structure] and made entry behind E11's crew and began to fight the fire (see Photo 5). The IC assigned BC6 to Division 3 (the second floor) and the crew from T11 to make entry at the front door to search the upper floors. When the T11 crew approached the door, RS1B had just come out. The two crews regrouped to go back in with BC6. E24 and RS1A backed out of the garage. RS1A proceeded to [the left side of the structure], E24 proceeded to the front door and met up with BC6,~T11,~and~RS1B.

At approximately 1105 hours, T15 radioed the IC that they were ready to vertically ventilate. The IC gave T15 the go ahead. At about that same time, RS1A made entry on [the left side of the structure] after the E11 crew knocked down the large room and contents fire. RSA1 radioed command that they were conducting a primary search. E24, T11 and RS1B regrouped at the front door, BC6 instructed T11 to stay at the front door and help feed the hoseline as BC6, E24 and RS1B made entry. After entering the front door, BC6, E24 and RS1B followed the E26 hoseline. BC6 had a thermal imaging camera, but the soot and smoke were so heavy he could not see the camera screen in the zero visibility. In zero visibility conditions, the E24 crew advanced to the bottom of the stairs of the first floor hallway. The E24 lieutenant followed E26's hoseline which led him to a metal door to the foyer which led to the garage and basement stairs. He pushed open the door and found [the lieutenant] in the first floor foyer at the top of the stairs to the basement. Then seconds later, E24's fire fighter/paramedic found [the fire fighter/paramedic] a few feet away in the first floor hallway just past the doorway to the foyer. Note: [The lieutenant] was found on his knees with his arms in a hunched over position in the foyer at the top of the basement stairs. [The fire fighter/paramedic] was found face down in the hallway toward the bedroom. The E26 hoseline was found in the foyer with [the lieutenant] and the door partially closed on it.

About this time dispatch contacted command and stated that an E20 fire fighter's radio emergency alarm bottom had been activated. Note: E20 was the replacement RIC and was en route. The IC didn't see E20 on scene and knew that they hadn't reported to command. The E24 crew could hear a PASS device going off. BC6 heard someone say we found a downed fire fighter and heard PASS alarms going off so he radioed a ``Mayday'' and reported that there was a downed fire fighter. Note: The Mayday by BC6 was overridden by dispatch's information about E20's emergency alarm. The lieutenant from RS1A came up the stairs from the fire floor and assisted the E24 lieutenant in carrying [the lieutenant] out the garage man door. While going along the garage wall and a vehicle, the [lieutenant's] low pressure SCBA hose and facepiece got caught underneath the vehicle's tire. RS1A's fire fighter cut the hose to expedite the extrication. The E24 fire fighter/paramedic moved [the fire fighter/paramedic] into the foyer doorway to the garage then the fire fighter/paramedic came out of garage and yelled, ``Need help, fire fighters down.'' The fire fighter from RS1A came up the stairs from the fire floor to the foyer and assisted a T11 fire fighter with carrying [the fire fighter/paramedic] outside through the garage. At approximately 1110 hours the IC radioed dispatch and urgently requested 2 medic units be sent; the IC requested a second alarm one minute later. The IC radioed BC6 and BC9 to report to the command post. The [lieutenant and the fire fighter/paramedic] were transported to the local medical center where the lieutenant was pronounced dead. The fire fighter/paramedic died two days later from his injuries.
\end{quote}

@@

\begin{table}[!ht]
\caption[Abridged NIOSH approximate fire event timeline]{Abridged NIOSH approximate fire event timeline~\cite{NIOSH:Bowyer2}}
\begin{tabular}{cl}
\toprule
Incident Time  &  Fire Behavior / Fireground Operation                       \\
{(hh:mm:ss)}   &                                                             \\
\midrule
17:15:00       &  Reported working fire with heavy smoke coming from the     \\ 
               &  front and rear of the structure's attic.                   \\
               &                                                             \\
17:16:00       &  Dispatch for a ``Still'' alarm - two engine companies,     \\
               &  two truck companies, and a battalion chief.                \\
               &                                                             \\
17:20:00       &  Attic windows break - heavy smoke out of front attic       \\
               &  window and heavy smoke and fire out of rear.               \\
               &                                                             \\
17:21:00       &  E123 captain and firefighter on second floor               \\
               &  with 1\sfrac{3}{4}-inch hoseline.                          \\
               &                                                             \\
17:21:30       &  First floor exterior porch door opened.                    \\
               &                                                             \\
17:23:00       &  Heavy fire report in rear stairwell and enclosed porch.    \\
               &  No response from E123 captain or firefighter.              \\
               &                                                             \\
17:24:00       &  Water applied to attic from the rear exterior position.    \\
               &                                                             \\
17:25:00       &  SQ5 makes entry through front of structure.                \\
               &                                                             \\
17:27:00       &  After hearing a Mayday call from SQ5, IC calls a Mayday.   \\
               &                                                             \\
17:29:00       &  E123 captain carried out of structure.                     \\
\bottomrule
\end{tabular}
\label{tab:incident_timeline}
\end{table}

>>>>

The account of events for this incident was documented by the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2011-13~\cite{NIOSH:Bowyer2}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer2}:
\begin{quote}
On June 02, 2011, a 48 year-old career lieutenant and a 53 year-old fire fighter/paramedic died in a multi-level residential structure fire while searching for the seat of the fire. Note: The residential structure where the fatalities occurred was built on a significantly sloped hillside common throughout the city. The fire floor was one floor below street level. Six companies and three command chiefs were dispatched to a report of an electrical fire at a residential home.

When Engine~26, staffed with a lieutenant, fire fighter/paramedic, and driver arrived at approximately 1048 hours, they noticed light smoke showing as they made entry through the front door, side A, street level, of the building. Minutes later, the incident commander (IC) tried contacting them over the radio, but received no response. A battalion chief (BC) assigned to ``the fire attack group'' followed the hoseline through the door and spoke to the [lieutenant and fire fighter/paramedic] on the street level floor. The lieutenant stated to the BC that the fire must be a floor below them. The BC stated they would attack the fire from the [left side] of the structure and exited the front door. The [lieutenant and fire fighter/paramedic] did not follow. A few minutes later the IC again tried to contact Engine~26 via radio with no response.

The crew from Engine~24, assigned to back up Engine~26, and a split crew from Rescue~1 tried to make entry through the door in the garage but could not advance due to the heat. The BC went to the [door on the left side of the structure], located one floor below street level, and forced the door with the Engine~11 crew on the hoseline. They immediately felt a blast of heat from the fully involved basement area. The Rescue~1 crew backed out of the garage and re-entered on [the left side of the structure] after the Engine~11 crew knocked down the large room and contents fire. At about the same time, the Engine~24 crew also backed out of the garage and followed the Engine~26 crew's hoseline through the front door. In zero visibility conditions, separate members of the Engine~24 crew independently found a downed member of the Engine~26 crew.

The Incident Commander was alerted of a downed fire fighter but, did not initially realize, until moments later that it was actually two downed fire fighters. Both [fire fighters] were removed from the structure and immediate medical treatment was provided. The [fire fighters] were transported to the local medical center where the lieutenant was pronounced dead and the fire fighter/paramedic died two days later.
\end{quote}

@@

\begin{table}[!ht]
\caption[Abridged NIOSH approximate fire event timeline]{Abridged NIOSH approximate fire event timeline~\cite{NIOSH:Bowyer2}}
\begin{tabular}{cl}
\toprule
Incident Time  &  Fire Behavior / Fireground Operation                                                \\
{(hh:mm:ss)}   &                                                                                      \\
\midrule
10:45:00       &  Dispatch for a curtain fire due to an electrical short at a residential structure.  \\
               &                                                                                      \\
10:48:00       &  E26 arrives on scene, reports light smoke conditions, and makes entry into          \\
               &  the front door with a 1 3/4'' hoseline.                                             \\
               &                                                                                      \\
10:54:00       &  E26 crew states that the fire is located below the first floor.                     \\
               &                                                                                      \\
10:56:00       &  BC9 observes smoke but no fire at the left rear corner of the structure.            \\
               &                                                                                      \\
10:58:00       &  Fire self-vents from the rear side of the structure when a glass door in the        \\
               &  basement fails.                                                                     \\
               &                                                                                      \\
10:59:00       &  BC9 forces open the basement door on the left side of the structure,                \\
               &  reports heavy fire and smoke, and requests a second hoseline.                       \\
               &                                                                                      \\
11:00:00       &  BC6 notices a severe change in conditions (heavy black smoke from garage).          \\
               &                                                                                      \\
11:01:00       &  Heavy fire and black smoke observed at rear of structure. Incident command          \\
               &  (IC) attempts to contact E26 several times via radio with no reply.                 \\
               &                                                                                      \\
11:02:00       &  E32 crew makes entry into the basement with a second hoseline.                      \\
               &                                                                                      \\
11:08:00       &  Two downed E26 firefighters are found on the first floor.                           \\
               &                                                                                      \\
11:09:00       &  The downed E26 firefighters are carried out of the structure.                       \\
               &                                                                                      \\
\bottomrule
\end{tabular}
\label{tab:incident_timeline}
\end{table}
